\chapter*{Acknowledgements}

To Hannah and Sam, I can't thank you enough for your patience, love, support, and encouragement, without which this thesis would never have been finished.
Please know that I am forever grateful for everything you've done.

To Dimitris and Liddy, thank you for your help and encouragement with my thesis.
I consider myself very lucky to have had two excellent and supportive supervisors.
I have learned so much that wouldn't have been possible without your guidance and encouragement.

To Mum and Dad, and Richard and Rosemary, thank you for making our move to Sheffield (and therefore this PhD) possible, and for all your support, encouragement, and help keeping us going over the last few years.

To Jo, thanks for encouraging me to start down this path.
It's been hard but your 'PhD survival kit' has helped me every step of the way.
The last pencil ran out just the other day.

To Tom, Amber, Martin, Becky, Azeezat, and everyone in D5, thanks for your sympathetic ear when I needed to moan, pizza when I needed to eat, and coffee when I needed to function.

Thanks to Laurie Mott of Doncaster Metropolitan Borough Council for providing advice, data, and context on Doncaster, and for the field research that I used to build the case studies.
Thank you to Emma Nicholas, also of DMBC and Well Doncaster, for showing me the community development projects in Denaby Main, one of my case study areas.

Thank you to Les Monaghan who kindly allowed me to reproduce the photographs in Chapter 1 from his 'Relative Poverty' project.

I would like to acknowledge the funding provided from Doncaster Metropolitan Borough Council, without which this project would not have been possible.
